neonstore: provenance-aware persistent storage for NEON data

### 2022-01-10 paper/paper.Rmd
:pen:

---
title: "neonstore: provenance-aware persistent storage for NEON data"
author:
  - name: "Carl Boettiger"
    affiliation: ucb
  - name: "Quinn Thomas"
    affiliation: vt
  - name: "Christine Laney"
    affiliation: neon
  - name: "Claire Lunch"
    affiliation: neon
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: neon
    address: "NEON"
  - code: vt
    address: "Virginia Tech"
abstract: |
  The National Ecological Observatory Network
journal: preprint
date: "`r Sys.Date()`"
#bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: rticles::elsevier_article
---

@@

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = FALSE
)
library(neonstore)
#Sys.setenv("NEONSTORE_HOME" = tempfile())
```

The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @NAS_GRAND].  
NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.  The `neonstore` R package seeks to provide provenance aware, high performance access and persistent storage of NEON data files. 
Here, we describe the design, use, and rationale for the package.
We illustrate a few example use cases, including downloading the 30 minute single-aspirated temperature data across all available NEON sites and dates, a collection of over 13 thousand individual files totalling nearly 2 gigabytes of data.
We demonstrate how `neonstore` facilitates summary calculations such as computing monthly means across all sites in a fraction of a second with minimal memory requirements.
The approaches taken in `neonstore` maximize reproducibility, facilitate collaboration including across computer languages, and help making analyses which leverage the full spatial and temporal scale of NEON easily available to researchers with modest computational resources and expertise.

@@

`neonstore` is also concerned with performance of research workflows based on NEON data.
Native operations in R are based in memory, making it difficult to work with objects that are larger than RAM.
Researchers seeking to take full advantage of NEON data collected across all available sites and years can quickly run ub against the limits of data which can be held and processed in working memory.
The R language also supports interfaces with relational databases [@dbi], with the widely used package `dplyr` capable of seamlessly tanslating common functions (e.g. `filter`, `select`, `mutate`, `group_by`, `summarise`, `inner_join`, etc) into SQL queries that can be run by external database servers [@dplyr; @dbplyr].
This allows users to perform common tasks on data much larger than would fit into working RAM.
Traditional relational databases such as MariaDB and Postgres use a server-client model with row-oriented data structures, suitable for concurrent read and write operations by many simultaneous client processes acccessing a central database (e.g. many client banks simultaneously updating a central registry of accounts, see ACID).
Modern data science workflows frequently benefit more from column-oriented data structures where simultaneous operations (ACID transactions) are not required.
This makes it possible to both eliminate the requirement of separately installing and running a database server while also opening the door to substantial performance improvements beyond the reach of traditional relational database clients.
`duckdb` is a leading example of such a columnar database [@duckdb], with seamless support for R integration and superior performance benchmarks to both traditional relational databases and in-memory operations in R using `dplyr`.

`neonstore` allows users to easily import tabular NEON data products (including `.h5` based eddy covariance products) into a local `duckdb` database using the function `neon_store()`.
If desired, users can instead use their own existing databases by providing the appropriate `DBI::dbConnection` object to the `db` argument of `neon_store()`. `neon_store()` stacks corresponding individual raw tabular data files by month, site, and (for IS data) sensor position into individual database tables.
`neon_store()` is capable of detecting revisions to previously released data products, removing outdated rows and replacing them with updated rows without accidental duplication of data.
Data revisions are indicated by downloading files which share the same data product, site, month and sensor position file name as files already present in the local store, but have an updated timestamp and differing content hash.
`neon_store()` provides a warning whenever existing data tables are updated, as this may require previous analyses to be re-run.
Because sensor data does not indicate the site or position information in individual data files (this is provided only through the file name) `neonstore` stacking functions add additional columns for these variables automatically.
Additionally, a `file` column is always included indicating the raw file source for each row.
Many NEON data products come in two types: a "basic" format and an "expanded" format which may include additional columns providing data fields that were not included in the original product specification.
NEON does not always use globally unique names for table descriptions, particularly for metadata tables such as `sensor_positions`. To avoid ambiguities created by these issues, database table names are formed using the table description name, the product type, and the product number, e.g. `brd_countdata-basic-DP1.10003.001`, rather than the description component alone, `brd_countdata`.

RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #1].

@@

`neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.

`neonstore` is distinguished as much by what it does not do as by what it does.
All `neonstore` operations are intended to be transparent and intuitive, and could easily be duplicated in another computational language or command line interface.
Specifically, `neonstore` will never automatically "clean" data: opaque data transformations make output sensitive to the version and implementation details of the tool in question.
A user should expect the same file from a "download" operation regardless of the software or software version used to perform the download.
Functions which in read in the data should faithfully reproduce the raw data so that analyses can easily be compared to those using other software tools.
When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behaviour of common functions used in the language.

@@

- `neonstore` aims to provide persistent storage, writing raw data files to
  the appropriate app directory for your operating system (see `rappdirs`,
  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)), now provided
  by the base R function `tools::R_user_dir()`.  Files are organized into appropriate 
  subdirectories by product, site, and year-month of sampling.  
- `neon_download()` provides  clean and concise progress bars for the two key
  processes involved: querying the API to obtain download URLs (which involves no 
  large data transfer but counts against API rate limiting, see below), and the
  actual file downloads.
- `neon_download()` will verify the integrity of file downloads against the MD5 or CRC32
  hashes provided. 
- `neon_download()` will omit downloads of any existing data files in the local 
  store (even if the timestamp portion of the filename has changed without updating the file)
- You can request multiple products at once using vector notation, though API
  rate limiting may interfere with large requests.
- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
- Automatic handling NEON's rate limiting policy.  To do, `neon_download()` must  pause 100 seconds every 150 requests if an API token is not supplied or every 950 requests if a token is available.  Should a rate-limit error be detected, (possible if the same token is used in parallel requests), `neon_download()` will also pause for 100 seconds before retrying the failed request.   Requests and pauses are communicated through the user messaging interface.  
- Expanding compressed HDF5 files (`*h5.gz`). (Note: unusually, NEON chooses to compress binary h5 files with gzip, even though gzip compression is built internally into the H5 specification at configurable compression levels.  NEON does not choose to compress text-based formats such as .csv, where compression has larger gains.  Compression has implications for version management, where `neonstore` must track the hashes of compressed files to avoid duplicate downloads)
- Updating the local manifest in LMDB with the individual checksum and current release tag for each downloaded object (see Provenance)

@@

For very large tables such as sensor data collections spanning many sites and years, users will require the increased performance provided by the high-performance, columnar-oriented relational database, `duckdb` [@duckdb].
To take advantage of this approach, we must first store the contents of the individual data files in the relational database using the function `neon_store()`:

@@

```{r}
neon_download(product = "DP1.00002.001", table = "30min") #Temp single aspirated
neon_store(product = "DP1.00002.001", table = "30min") #Temp single aspirated
```

@@

`neonstore`'s core dependencies are quite light, avoiding hard requirements on other packages which themselves bring in a large recursive list of dependencies. `DBI` [@DBI], `duckdb` [@duckdb], and `thor` [@thor] provide the core database interfaces; all have no recursive dependencies. `httr` [@httr] provides functionality required to interface with the NEON REST API, and `R.utils`, `zip`, and `vroom` [@vroom] provide parsers required to unpack the data. `progress` provides user-friendly progress bars and estimated run-times for slow functions.

Optional dependencies are required for specific tasks. `dplyr` is needed for lazy read in `neon_table()`.
`neonUtilities` [@neonUtilities] and `rhdf5` are required at this time for reading in eddy-covariance products.
`arrow` [@arrow] is required for remote database connections using `neon_remote()`.
`rstudioapi` is required to utilize `neon_pane()` to view the connections pane from within RStudio.
The remaining suggested dependencies are utilized within testing only.

## Appendix

Discuss the NEON filename convention parser, link to `neon_filename_parser.R` source for regex patterns.  
Discuss cases of NEON changing filename conventions.

>>>>

---
title: "neonstore: provenance-aware persistent storage for NEON data"
author:
  - name: "Carl Boettiger"
    affiliation: ucb
  - name: "R. Quinn Thomas"
    affiliation: vt
  - name: "Christine Laney"
    affiliation: neon
  - name: "Claire Lunch"
    affiliation: neon
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: neon
    address: "NEON"
  - code: vt
    address: "Virginia Tech"
abstract: |
  The National Ecological Observatory Network
journal: preprint
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: rticles::elsevier_article
---

@@

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  cache = FALSE
)
library(neonstore)
Sys.setenv("NEONSTORE_HOME" = "store")
```

The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @Balch2020; @Nagy2021].
NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.
The `neonstore` R package seeks to provide provenance aware, high performance access and persistent storage of NEON data files. 
Here, we describe the design, use, and rationale for the package.
We illustrate a few example use cases, including downloading the 30 minute single-aspirated temperature data across all available NEON sites and dates, a collection of over 13 thousand individual files totalling nearly 2 gigabytes of data.
We demonstrate how `neonstore` facilitates summary calculations such as computing monthly means across all sites in a fraction of a second with minimal memory requirements.
The approaches taken in `neonstore` maximize reproducibility, facilitate collaboration including across computer languages, and help making analyses which leverage the full spatial and temporal scale of NEON easily available to researchers with modest computational resources and expertise.

@@

`neonstore` is also concerned with performance of research workflows based on NEON data.
Native operations in R are based in memory, making it difficult to work with objects that are larger than RAM.
Researchers seeking to take full advantage of NEON data collected across all available sites and years can quickly run ub against the limits of data which can be held and processed in working memory.
The R language also supports interfaces with relational databases [@dbi], with the widely used package `dplyr` capable of seamlessly tanslating common functions (e.g. `filter`, `select`, `mutate`, `group_by`, `summarise`, `inner_join`, etc) into SQL queries that can be run by external database servers [@dplyr; @dbplyr].
This allows users to perform common tasks on data much larger than would fit into working RAM.
Traditional relational databases such as MariaDB and Postgres use a server-client model with row-oriented data structures, suitable for concurrent read and write operations by many simultaneous client processes acccessing a central database (e.g. many client banks simultaneously updating a central registry of accounts, see ACID).
Modern data science workflows frequently benefit more from column-oriented data structures where simultaneous operations (ACID transactions) are not required.
This makes it possible to both eliminate the requirement of separately installing and running a database server while also opening the door to substantial performance improvements beyond the reach of traditional relational database clients.
`duckdb` is a leading example of such a columnar database [@Raasveldt_2019; @duckdb], with seamless support for R integration and superior performance benchmarks to both traditional relational databases and in-memory operations in R using `dplyr`.

`neonstore` allows users to easily import tabular NEON data products (including `.h5` based eddy covariance products) into a local `duckdb` database using the function `neon_store()`.
If desired, users can instead use their own existing databases by providing the appropriate `DBI::dbConnection` object to the `db` argument of `neon_store()`. `neon_store()` stacks corresponding individual raw tabular data files by month, site, and (for IS data) sensor position into individual database tables.
`neon_store()` is capable of detecting revisions to previously released data products, removing outdated rows and replacing them with updated rows without accidental duplication of data.
Data revisions are indicated by downloading files which share the same data product, site, month and sensor position file name as files already present in the local store, but have an updated timestamp and differing content hash.
`neon_store()` provides a warning whenever existing data tables are updated, as this may require previous analyses to be re-run.
Because sensor data does not indicate the site or position information in individual data files (this is provided only through the file name) `neonstore` stacking functions add additional columns for these variables automatically.
Additionally, a `file` column is always included indicating the raw file source for each row.
Many NEON data products come in two types: a "basic" format and an "expanded" format which may include additional columns providing data fields that were not included in the original product specification.
NEON does not always use globally unique names for table descriptions, particularly for metadata tables such as `sensor_positions`. To avoid ambiguities created by these issues, database table names are formed using the table description name, the product type, and the product number, e.g. `brd_countdata-basic-DP1.10003.001`, rather than the description component alone, `brd_countdata`.
RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #1].

![RStudio Connections pane showing interactive panel listing all tables imported into the local duckdb database.](rstudio-pane.png)

<!--

@@

-->

`neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.
`neonstore` is distinguished as much by what it does not do as by what it does.
All `neonstore` operations are intended to be transparent and intuitive, and could easily be duplicated in another computational language or command line interface.
Specifically, `neonstore` will never automatically "clean" data: opaque data transformations make output sensitive to the version and implementation details of the tool in question.
A user should expect the same file from a "download" operation regardless of the software or software version used to perform the download.
Functions which in read in the data should faithfully reproduce the raw data so that analyses can easily be compared to those using other software tools.
When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behaviour of common functions used in the language.

@@

- `neonstore` aims to provide persistent storage, writing raw data files to
  the appropriate app directory for your operating system (see `rappdirs`,
  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)), now provided by the base R function `tools::R_user_dir()`.
  Files are organized into appropriate subdirectories by product, site, and year-month of sampling.
- `neon_download()` provides  clean and concise progress bars for the two key
  processes involved: querying the API to obtain download URLs (which involves no 
  large data transfer but counts against API rate limiting, see below), and the
  actual file downloads.
- `neon_download()` will verify the integrity of file downloads against the MD5 or CRC32
  hashes provided. 
- `neon_download()` will omit downloads of any existing data files in the local 
  store (even if the timestamp portion of the filename has changed without updating the file)
- You can request multiple products at once using vector notation, though API
  rate limiting may interfere with large requests.
- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
- Automatic handling NEON's rate limiting policy.
  To do this, `neon_download()` must  pause 100 seconds every 150 requests if an API token is not supplied or every 950 requests if a token is available.
  Should a rate-limit error be detected, (possible if the same token is used in parallel requests), `neon_download()` will also pause for 100 seconds before retrying the failed request.
  Requests and pauses are communicated through the user messaging interface.
- Expanding compressed HDF5 files (`*h5.gz`).
  (Note: unusually, NEON chooses to compress binary h5 files with gzip, even though gzip compression is built internally into the H5 specification at configurable compression levels.
  NEON does not choose to compress text-based formats such as .csv, where compression has larger gains.
  Compression has implications for version management, where `neonstore` must track the hashes of compressed files to avoid duplicate downloads)
- Updating the local manifest in LMDB with the individual checksum and current release tag for each downloaded object (see Provenance)

@@

For very large tables such as sensor data collections spanning many sites and years, users will require the increased performance provided by the high-performance, columnar-oriented relational database, `duckdb` [@Raasveldt_2019; @duckdb].
To take advantage of this approach, we must first store the contents of the individual data files in the relational database using the function `neon_store()`:

@@

```{r download_temp}
neon_download(product = "DP1.00002.001", table = "30min") #Temp single aspirated
neon_store(product = "DP1.00002.001", table = "30min") #Temp single aspirated
```

@@

`neonstore`'s core dependencies are quite light, avoiding hard requirements on other packages which themselves bring in a large recursive list of dependencies. `DBI` [@dbi], `duckdb` [@duckdb], and `thor` [@thor] provide the core database interfaces; all have no recursive dependencies.
`httr` [@httr] provides functionality required to interface with the NEON REST API, and `R.utils`, `zip`, and `vroom` [@vroom] provide parsers required to unpack the data.
`progress` provides user-friendly progress bars and estimated run-times for slow functions.
Optional dependencies are required for specific tasks. `dplyr` is needed for lazy read in `neon_table()`.
`neonUtilities` [@neonUtilities] and `rhdf5` are required at this time for reading in eddy-covariance products.
`arrow` [@arrow] is required for remote database connections using `neon_remote()`.
`rstudioapi` is required to utilize `neon_pane()` to view the connections pane from within RStudio.
The remaining suggested dependencies are utilized within testing only.

## Conclusions

`neonstore` provides a high-performance and provenance-aware mechanism for accessing, maintaining, and working with NEON data products, designed to scale seamlessly to analyses which stretch across the full spatial and temporal range which makes this nation-wide dataset so uniquely valuable. 
The `neonstore` package follows simple and robust best-practices for data management that can work across most popular computing languages; such as preserving a local library of the original raw data files downloaded from NEON, and importing data into a local relational database (`duckdb`) that can be accessed from a wide range of libraries.
Additional utilities enable serverless network-based sharing of a large `neonstore` database across a team through the use of `parquet` and `arrow`'s remote filesystems (which likewise supports most major computing languages.)
`neonstore` has been co-developed by NEON staff (CL & CL) and members of the Ecological Forecasting Initiative (CB & RQT), tested in supporting the EFI NEON forecasting challenge [@efi_challenge], among other use cases.

# References
